\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{conjecture}[theorem]{Conjecture}

\newcommand{\zD}{\zeta_D}
\newcommand{\zbar}{\bar\zeta}

\title{Turning Euler products into ratios of zeta values\\
\large Working notes --- with Mathar (arXiv:1106.4038) as prior art}
\author{}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Start with a multiplicative arithmetic function $a(n)$ and ask for a usable number
for $\sum_n a(n)/n^s$. Multiplicativity turns the sum into an Euler product, and if
the local factor is a rational function of $q:=p^{-1}$ we can peel it apart, one
order at a time, into pieces each of which is a Riemann zeta value. Throughout we
work in $q=p^{-1}$, so the Dirichlet exponent $s$ stays visible in the powers of
$q$ (and the exponent on $q$ is exactly the $\zeta$-argument). These notes tell the
story: the identity that makes it work, the little greedy algorithm that does the
peeling, when it stops (a clean closed form) versus when it does not (an honest
infinite product), a cheap way to compute it in $\log$-space, and the running
example $\sum_n 1/(n\varphi(n))$ together with the harder target $1/\varphi(n)^2$.
R.~J.~Mathar's survey \cite{Mathar2011} turns out to set up exactly the same
construction, so we use it to calibrate before pushing further.
\end{abstract}

\tableofcontents

\section{A concrete question to anchor everything}
\label{sec:anchor}

The first thing worth doing is picking a number to chase. Take
\[
\sum_{n\ge1}\frac{1}{n\,\varphi(n)},
\]
where $\varphi$ is Euler's totient. Since $n\mapsto 1/(n\varphi(n))$ is
multiplicative, the sum factors over primes,
\[
\sum_{n\ge1}\frac{1}{n\varphi(n)}
=\prod_{p}\sum_{\alpha\ge0}\frac{1}{p^{\alpha}\varphi(p^{\alpha})}
=\prod_{p}\Big(1+\sum_{\alpha\ge1}\frac{1}{p^{2\alpha}(1-p^{-1})}\Big),
\]
the split in the last step being there because $\varphi(p^\alpha)=p^\alpha(1-p^{-1})$
only holds for $\alpha\ge1$. The inner sum is geometric, and after clearing
denominators the local factor collapses to a tidy rational function of $q:=p^{-1}$,
\[
\sum_{n\ge1}\frac{1}{n\varphi(n)}
=\prod_p\frac{1-q+q^{3}}{1-q-q^{2}+q^{3}},\qquad q=p^{-1}.
\]
Now notice the thing that makes the whole method go: the numerator and denominator
are \emph{almost} equal --- they differ only by the single term $q^2$. If we could
nudge that mismatch up into higher and higher powers of $q$, the leftover would
matter less and less, and we would be left with an approximation we can actually
evaluate. That is the whole idea, and it works because we happen to know, in closed
form, the product over primes of the nudging factors.

\subsection{The two moves, and why they are zeta values}

The moves we are allowed are ``multiply and divide by $(1\mp q^{a})$,'' and the
reason we like them is Euler's product for $\zeta$. With $q=p^{-1}$ (so
$q^{s}=p^{-s}$),
\begin{align}
\zeta(s) &= \prod_p\frac{1}{1-q^{s}}, &
\zbar(s) &:= \frac{\zeta(2s)}{\zeta(s)} = \prod_p\frac{1}{1+q^{s}}. \label{eq:zzbar}
\end{align}
So every factor $\prod_p(1-q^{a})^{\pm1}$ or $\prod_p(1+q^{a})^{\pm1}$ we introduce
simply accumulates a $\zeta(a)$ or a $\zbar(a)$ --- the exponent $a$ on $q$ is the
$\zeta$-argument. For instance, multiplying and dividing the $1/(n\varphi(n))$
factor by $(1+q^2)$ shifts the denominator to $1-q-q^4+q^5$, pushing the mismatch
from order $q^2$ up to order $q^3$, while the $(1+q^2)$ we introduced contributes
exactly a $\zbar(2)^{-1}$. Repeating this bookkeeping is the whole game; run it nine
times and you already get
\[
\sum_{n\ge1}\frac{1}{n\varphi(n)}
\;\approx\;
\frac{\zeta(2)\zeta(3)\zeta(4)\zeta(5)\zeta(13)\zeta(14)\zeta(16)\zeta(18)\zeta(20)}
     {\zeta(8)\zeta(9)\zeta(10)}
\;=\;2.203927707\ldots
\]
The rest of these notes is about making this precise, deciding when it terminates,
and computing it without drowning in algebra.

\section{The general setup}
\label{sec:setup}

\begin{definition}[Dirichlet series]
For an arithmetic function $a:\mathbb{Z}_{\ge1}\to\mathbb{C}$,
\[
\zD(s) \;\equiv\; \sum_{n\ge1} \frac{a(n)}{n^s}.
\]
\end{definition}

When $a$ is multiplicative, $\zD$ factors over primes into local (Bell) factors
\[
\zD(s) = \prod_p B_p, \qquad B_p \equiv \sum_{e\ge0} a(p^e)\, q^{se}, \qquad q=p^{-1}.
\]
For every function we care about, $B_p$ is a \emph{rational} function of $q$ (with
$s$ living in the exponents), whose coefficients are polynomial or rational
functions of $p$. That rationality --- more than multiplicativity itself --- is what
lets the extraction go through: we can always write
$B_p=\mathrm{NUM}_p(q)/\mathrm{DEN}_p(q)$, normalized so that $B_p\to1$ as $q\to0$.

\subsection{The identity that makes it work}

The two moves of \S\ref{sec:anchor} generalize to any shift. Write the shift
argument $a=us-l$ with $u\in\mathbb{Z}_{\ge1}$, $l\in\mathbb{Z}$ (so
$q^{a}=q^{us-l}=p^{\,l-us}$). Then for every exponent $\gamma\in\mathbb{Z}$,
\begin{align}
\prod_p (1 - q^{\,a})^{\gamma} &= \zeta(a)^{-\gamma}, \label{eq:euler-minus}\\
\prod_p (1 + q^{\,a})^{\gamma} &= \frac{\zeta(a)^{\gamma}}{\zeta(2a)^{\gamma}}. \label{eq:euler-plus}
\end{align}
Equation~\eqref{eq:euler-minus} is just \eqref{eq:zzbar} for $\zeta$ with the
argument shifted; \eqref{eq:euler-plus} follows by writing $1+y=(1-y^2)/(1-y)$ with
$y=q^{a}$ and applying \eqref{eq:euler-minus} twice. (The special case $a=s$, i.e.\
$u=1,l=0$, is precisely $\zeta$ and $\zbar$ from before.) The point to keep in mind:
\emph{the exponent on $q$ is the $\zeta$-argument, and the $1+$ branch doubles it}.

So the extraction problem is this: given $B_p=\mathrm{NUM}_p(q)/\mathrm{DEN}_p(q)$,
find a (possibly infinite) factorization
\begin{equation}
\label{eq:master-factorization}
B_p = \prod_{i\ge1} (1 - S_i\, q^{\,u_i s - l_i})^{\gamma_i}, \qquad
S_i\in\{+1,-1\},\ l_i,u_i,\gamma_i\in\mathbb{Z},\ u_i\ge1,
\end{equation}
ordered by increasing $u_i$ (and, for ties, decreasing $l_i$). Feeding each factor
through \eqref{eq:euler-minus}--\eqref{eq:euler-plus} then reads off $\zD(s)$ as a
product/ratio of zeta values.

\subsection{The algorithm, told as a story}
\label{sec:algo}

The peeling is greedy and almost embarrassingly simple:
\begin{enumerate}
\item Expand $\mathrm{NUM}_p$ and $\mathrm{DEN}_p$ as power series in $q$ (both with
constant term $1$) up to a working order $N$.
\item Find the lowest order at which they still disagree. The size of that
disagreement pins down exactly one binomial factor $(1-S\,q^{a})^{\gamma}$ whose own
expansion cancels the discrepancy at that order.
\item Fold that factor in (multiply it into the numerator, record its zeta value)
and go back to step 2 on the new residual.
\item If the residual ever becomes exactly $1$, stop --- you have a finite closed
form. Otherwise keep going; the exponents $(l_i,u_i,\gamma_i)$ form a sequence you
try to recognize.
\end{enumerate}
Each pass raises the order of agreement by at least one, so as a formal power series
this always converges; the only question is whether it ever \emph{halts}
(\S\ref{sec:termination}).

Two established procedures are hiding in here. \textbf{Selberg--Delange} writes
$F(s)=\zeta(s)^{\rho}G(s)$ with $G$ holomorphic and non-vanishing near $s=1$ and
needs only finitely many derivatives of $G$; we are being greedier, asking to
represent $G$ \emph{itself} as an infinite product of zeta values. And peeling off
the next-order term and recursing is exactly the \textbf{plethystic logarithm} of
the physics literature (Feng--Hanany--He); it terminates in the same ``nice'' cases
and runs forever in general.

\section{Mathar got here first}

R.~J.~Mathar's \emph{Survey of Dirichlet Series of Multiplicative Arithmetic
Functions} \cite{Mathar2011} sets up \emph{precisely} the identity of
\S\ref{sec:setup} (his eq.~(1.5)--(1.7)): the same infinite-product form
\eqref{eq:master-factorization}, the same ordering convention (smallest $u_i$ first,
ties broken by largest $l_i$), and the same $\zeta/\zbar$ substitution. He is also
explicit that truncating the product is a numerical tool for $\zD(s)$, and that in
the nice cases the prime product collapses to a finite ratio of cyclotomic
polynomials in $q$ --- hence a finite product of zeta values.

This does not make the project redundant. Mathar's paper is a table of results with
sketched derivations, not a general algorithm with a reusable implementation, and he
never touches the plethystic / Selberg--Delange framing. But it is exactly the right
thing to (a) validate the extractor against and (b) cite as prior art.

\subsection{A worked check: the Hadamard square $\varphi^2(n)$}

Mathar's \S3.13.1 works the case most useful to us as a test. The totient has master
equation $\varphi(p^e)=(p-1)p^{e-1}$ ($e>0$) and the finite (terminating) Bell
series $\varphi(n)/n^s\mapsto\zeta(s-1)/\zeta(s)$. Squaring $\varphi$ pointwise (a
Hadamard square, \emph{not} Dirichlet self-convolution) gives $a(p^e)=(p-1)^2p^{2e-2}$
and
\[
B_p=\frac{1-2q^{\,s-1}+q^{\,s}}{1-q^{\,s-2}}\qquad\text{(his eq.~(3.78))}.
\]
The numerator is \emph{not} a clean binomial power --- $1-2q^{s-1}+q^{s}$ is not
$1-S q^{a}$ for a single shift --- so the extraction cannot stop at one factor, and
Mathar reports an infinite product (his eq.~(3.79)):
\begin{align}
\frac{\varphi^2(n)}{n^s} &\mapsto \zeta(s-2)\prod_p
(1-q^{s-1})^2(1+q^{s})^{-1}(1-q^{2s-2})^{-1}(1+q^{2s-1})^{-2}\notag\\
&\quad\times(1-q^{3s-3})^{-2}(1+q^{3s-2})^{-4}(1-q^{3s-1})^{-2}
(1-q^{4s-4})^{-3}(1+q^{4s-3})^{-8}(1-q^{4s-2})^{-5}\notag\\
&\quad\times(1+q^{4s-1})^{-2}(1-q^{5s-5})^{-6}(1+q^{5s-4})^{-16}(1-q^{5s-3})^{-16}
(1+q^{5s-2})^{-8}(1-q^{5s-1})^{-2}\cdots \label{eq:mathar-phi2}
\end{align}
The exponents $2,1,1,2,2,4,2,3,8,5,2,6,16,16,8,2,\dots$ are exactly the sort of
sequence our extractor should reproduce. With 16 published terms, this is a fully
independent check to run before trusting the code on anything new.

\begin{remark}
$\varphi^2(n)$ (pointwise square) is \emph{not} our eventual target
$1/\varphi(n)^2$ (pointwise reciprocal square); they are related only in that both
resist a clean closed form. We set up our own case in \S\ref{sec:target}.
\end{remark}

\section{When does it stop?}
\label{sec:termination}

The extraction always converges as a formal series, so the interesting question is
when it \emph{halts} --- when the residual hits exactly $1$ after finitely many
factors. When it does, the output is a finite product
\[
B_p = \prod_{i=1}^{M} (1 - S_i\, q^{\,u_i s - l_i})^{\gamma_i},
\]
so ``it terminates'' is almost tautologically ``$B_p$ \emph{is} such a finite
product.'' The useful content is recognizing that directly from
$\mathrm{NUM}_p,\mathrm{DEN}_p$, without running anything.

\begin{definition}[Admissible binomial]
An \emph{admissible binomial} is a factor $1-S\,q^{\,us-l}$ with $S\in\{\pm1\}$,
$u\ge1$, $l\in\mathbb Z$ --- a single power of $q$ whose exponent $a=us-l$ is the
$\zeta$-argument it will become.
\end{definition}

\begin{theorem}[Termination criterion]
\label{thm:termination}
Normalize $B_p=\mathrm{NUM}_p(q)/\mathrm{DEN}_p(q)$ so $B_p\to1$ as $q\to0$. The
extraction terminates if and only if both $\mathrm{NUM}_p$ and $\mathrm{DEN}_p$
factor into admissible binomials.
\end{theorem}

\begin{proof}[Sketch]
\emph{Sufficiency.} If $B_p$ is a finite product of admissible binomials, order them
by increasing $u_i$ (ties by decreasing $l_i$). The leading term of the smallest
surviving factor is exactly the lowest-order discrepancy the greedy step sees, so
each pass cancels precisely one factor and recurses on a strictly shorter product,
halting after $M$ passes. \emph{Necessity.} If the algorithm halts, the emitted
factorization is finite by definition, and reading it factor by factor is the
statement.
\end{proof}

\subsection{The numeric (single-variable) case}

When the arguments carry no power of $p$ --- all $l_i=0$, which is the situation of
$\sum_n 1/(n\varphi(n))$, where $B_p$ is a rational function of $q=p^{-1}$ with
$p$-free coefficients --- the admissible binomials are just $1-q^u$ and
$1+q^u=(1-q^{2u})/(1-q^u)$. Since $1-q^n=\prod_{d\mid n}\Phi_d(q)$ with $\Phi_d$ the
$d$-th cyclotomic polynomial, and
\[
\prod_n (1-q^n)^{m_n}=\prod_d \Phi_d(q)^{e_d},\qquad e_d=\sum_{n:\,d\mid n} m_n,
\]
is a bijection between finitely supported integer sequences (M\"obius inversion on
divisibility, $m_n=\sum_{n\mid k}\mu(k/n)e_k$), Theorem~\ref{thm:termination}
specializes cleanly:

\begin{proposition}
\label{prop:cyclotomic}
For $B_p$ with $p$-free coefficients and $B_p\to1$ as $q\to0$, the extraction
terminates iff $\mathrm{NUM}_p$ and $\mathrm{DEN}_p$ are products of cyclotomic
polynomials in $q$ --- equivalently iff every zero and pole of $B_p$ is a root of
unity --- and the value is then the finite monomial
$\prod_p B_p=\prod_n\zeta(n)^{-m_n}$.
\end{proposition}

\subsection{Where it terminates (and it usually does)}

Termination is in fact generic among the classical multiplicative functions: any
Bell factor that is a ratio of admissible binomials halts, and Mathar's totient
$\varphi(n)/n^s\mapsto\zeta(s-1)/\zeta(s)$ is just the prototype. All of the
following stop in one or two factors (with $q=p^{-1}$, so the exponent on $q$ is the
$\zeta$-argument):

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{lll}
\hline
summand & $B_p$ & $\zD(s)$ \\
\hline
$1/n^s$ & $(1-q^{s})^{-1}$ & $\zeta(s)$ \\
$n^k/n^s$ \ ($\mathrm{Id}_k$) & $(1-q^{s-k})^{-1}$ & $\zeta(s-k)$ \\
$\mu(n)/n^s$ & $1-q^{s}$ & $1/\zeta(s)$ \\
$\lambda(n)/n^s$ \ (Liouville) & $(1+q^{s})^{-1}$ & $\zeta(2s)/\zeta(s)$ \\
$\mu^2(n)/n^s$ \ (squarefree ind.) & $1+q^{s}$ & $\zeta(s)/\zeta(2s)$ \\
$\varphi(n)/n^s$ & $\dfrac{1-q^{s}}{1-q^{s-1}}$ & $\zeta(s-1)/\zeta(s)$ \quad[Mathar (3.76)]\\
$J_k(n)/n^s$ \ (Jordan) & $\dfrac{1-q^{s}}{1-q^{s-k}}$ & $\zeta(s-k)/\zeta(s)$ \quad[Mathar (3.87)]\\
$\sigma_k(n)/n^s$ & $\big[(1-q^{s})(1-q^{s-k})\big]^{-1}$ & $\zeta(s)\,\zeta(s-k)$ \\
$d(n)/n^s$ \ ($\sigma_0$) & $(1-q^{s})^{-2}$ & $\zeta(s)^2$ \\
$2^{\omega(n)}/n^s$ & $\dfrac{1+q^{s}}{1-q^{s}}$ & $\zeta(s)^2/\zeta(2s)$ \\
\hline
\end{tabular}
\end{center}

The last row is a clean illustration of the $1+q^{s}$ (i.e.\ $\zbar$) branch:
$\prod_p(1+q^{s})=\zeta(s)/\zeta(2s)$ and $\prod_p(1-q^{s})^{-1}=\zeta(s)$ combine to
$\zeta(s)^2/\zeta(2s)$.

\subsection{Where it fails --- and it is always the numerator's fault}

The non-terminating cases all fail Theorem~\ref{thm:termination} in the
\emph{numerator}:
\begin{itemize}
\item \textbf{$\varphi^2(n)$ (Mathar \S3.13.1).}
$B_p=\dfrac{1-2q^{s-1}+q^{s}}{1-q^{s-2}}$: the denominator $1-q^{s-2}$ is admissible,
but the numerator $1-2q^{s-1}+q^{s}$ is not a single admissible binomial, and the
product is genuinely infinite \eqref{eq:mathar-phi2}.
\item \textbf{$1/\varphi(n)^2$ (our target).} By
Proposition~\ref{prop:bell-1overphi2} the numerator coefficient
$\frac{2p-1}{(p-1)^2p^2}$ is not a monomial in $p$; same obstruction.
\item \textbf{$\sum_n 1/(n\varphi(n))$ (numeric).} Here
$B_p=\dfrac{1-q+q^3}{1-q-q^2+q^3}$ with $q=p^{-1}$. The denominator factors as
$(1-q)(1-q^2)$, but the numerator $1-q+q^3$ has a real zero $q\approx-1.3247$ off
the unit circle, so by Proposition~\ref{prop:cyclotomic} the extraction cannot
terminate --- which is why we only ever use it for the truncated approximation
$\approx 2.2039$ of \S\ref{sec:anchor}.
\end{itemize}

\begin{remark}[It is structurally the numerator]
In every failure the denominator is already a product of admissible
binomials/cyclotomics; termination breaks solely on the numerator. That is not an
accident: denominators come from geometric sums $\sum_e q^{(s-k)e}$ and are
automatically admissible, whereas the numerator encodes the arithmetic ``defect'' of
$a(p^e)$ and need not be.
\end{remark}

\begin{remark}[A sign convention that can defeat an otherwise finite case]
The FORM/Mathematica drivers keep the extracted exponents nonnegative, switching to
the $1+q^{a}$ branch (via \eqref{eq:euler-plus}) whenever the leading discrepancy is
negative. Because $1+q^{a}=(1-q^{2a})/(1-q^{a})$ this stays admissible but does not
peel a $1-q^{a}$ factor in reverse: on input $B_p=1-q$ the greedy runs
$1-q\to1-q^2\to1-q^4\to\cdots$ forever, doubling the order, even though $B_p$ is a
single admissible binomial. To make the driver's halting match
Theorem~\ref{thm:termination}, the negative-coefficient branch must be allowed to
emit $(1-q^{a})^{\gamma}$ with $\gamma<0$ (accumulating $\zeta(a)^{\gamma}$) rather
than routing through $1+q^{a}$.
\end{remark}

\section{Computing it without blowing up: the logarithmic stream}
\label{sec:logstream}

There is a naive way to run the algorithm and a much better one. The naive way keeps
the numerator as a fully expanded polynomial and multiplies each new binomial into
it; the degree balloons and the whole thing chokes (our FORM driver does exactly
this, and hangs around order 50). The better way works in $\log$-space, where
\emph{multiplying} binomials becomes \emph{adding} their logarithms --- and those
logarithms are sparse.

Fix the numeric case: let $\mathrm{Num}(q)$ and $\mathrm{Den}(q)$ be integer
polynomials with constant term $1$, and
\[
B(q)=\frac{\mathrm{Num}(q)}{\mathrm{Den}(q)}=1+\sum_{k\ge1}B_k q^k .
\]
Because $\mathrm{Den}(0)=1$, its inverse is again an integer series, so every
$B_k\in\mathbb{Z}$. Write $[q^k]f$ for the coefficient of $q^k$ in $f$.

\subsection{Two ways to carry the state}

\paragraph{Direct.} Keep the integer coefficients of the running numerator $N$
(initially $\mathrm{Num}$). The residual at order $k$ is
$c=[q^k]N-[q^k]\mathrm{Den}$; folding a factor \emph{convolves} $N$ with a binomial.

\paragraph{Log.} Keep instead the residual logarithm
$R(q)=\log N(q)-\log\mathrm{Den}(q)$ (initially $\log B$). The only atoms we ever add
are sparse --- supported on multiples of $k$:
\begin{align}
\log(1-q^k) &= -\sum_{m\ge1}\tfrac{1}{m}q^{km}
            = -\big(q^k+\tfrac12 q^{2k}+\tfrac13 q^{3k}+\cdots\big),\label{eq:logminus}\\
\log(1+q^k) &= \sum_{m\ge1}\tfrac{(-1)^{m-1}}{m}q^{km}
            = q^k-\tfrac12 q^{2k}+\tfrac13 q^{3k}-\cdots.\label{eq:logplus}
\end{align}
Folding a factor now just \emph{adds} a multiple of \eqref{eq:logminus} or
\eqref{eq:logplus} to $R$ --- an addition of a sparse series instead of a
convolution. The one catch is the $\tfrac1m$: $R$ carries rational coefficients, and
you might worry that ``reading off'' the next factor is no longer a clean integer.
It is.

\subsection{Why the read is always an integer}

Sweep $k=1,2,3,\dots$, cancelling as you go, and read $c=[q^k]R$ once all lower
orders are zero.

\begin{lemma}[Integrality at the reading order]
\label{lem:integer}
If $[q^j]R=0$ for all $0<j<k$, then $c=[q^k]R\in\mathbb{Z}$, and in fact
$c=[q^k](N/\mathrm{Den})$ --- exactly the integer the direct method reads.
\end{lemma}

\begin{proof}
Let $S=N/\mathrm{Den}$, so $R=\log S$ and $S=\exp(R)$. With $[q^j]R=0$ for $0<j<k$,
\[
S=\exp(R)=1+[q^k]R\,q^k+O(q^{k+1}),
\]
since every nonlinear term of $\exp$ is a product of two or more terms of $R$, each
of order $\ge k$, hence of order $\ge 2k>k$. So $[q^k]S=[q^k]R$. But $N$ and
$\mathrm{Den}^{-1}$ are integer series, so $S$ is too, and $[q^k]S\in\mathbb{Z}$.
\end{proof}

\begin{remark}[Where the fractions hide]
The $\tfrac1m$ parts of \eqref{eq:logminus}--\eqref{eq:logplus} really do land in
$R$ --- but only at orders \emph{above} the current $k$ (the first fractional term
of $\log(1\pm q^k)$ is $\mp\tfrac12 q^{2k}$). By the time the sweep reaches any
order, everything below it has been cancelled, so Lemma~\ref{lem:integer} applies
afresh: you never actually read a fraction.
\end{remark}

\subsection{Choosing the factor, and the zeta it becomes}

With $c=[q^k]R$ in hand as an integer, use the same branch as the direct method:
\[
c>0:\ f=(1-q^k)^{c},\qquad c<0:\ f=(1+q^k)^{-c},
\]
and add $\log f$ to $R$. By \eqref{eq:logminus}--\eqref{eq:logplus} either choice has
$[q^k]\log f=-c$, so $R$'s order-$k$ coefficient goes to $0$ and only higher orders
move. In the single-$q$ toy $\zeta(k)\leftrightarrow 1/(1-q^k)$ and
$1+q^k=\zeta(k)/\zeta(2k)$, so each folded factor contributes to $B$
\[
c>0:\ \zeta(k)^{c}, \qquad c<0:\ \zeta(k)^{c}\,\zeta(2k)^{-c},
\]
with exponents accumulated as \emph{net} totals per argument --- so a later step can
cancel a $\zeta$ emitted earlier (e.g.\ a $1+q^8$ emits $\zeta(16)$, and a later
$1+q^{16}$ removes it).

\subsection{A worked micro-example}

Take $\mathrm{Num}=1-q+q^3$, $\mathrm{Den}=1-q-q^2+q^3$ again. To fourth order,
\[
\log\mathrm{Num}=-q-\tfrac12 q^2+\tfrac23 q^3+\tfrac34 q^4+\cdots,\qquad
\log\mathrm{Den}=-q-\tfrac32 q^2-\tfrac13 q^3-\tfrac34 q^4+\cdots,
\]
so their difference is \emph{not} integral,
\[
L=\log\mathrm{Num}-\log\mathrm{Den}
 =q^2+q^3+\tfrac32 q^4+\cdots .
\]
Order $1$ cancels; at order $2$ the halves and three-halves combine to the integer
$[q^2]L=1$ --- the first factor (power $2$, coefficient $+1$). But at order $4$ the
raw coefficient is the \emph{fraction} $\tfrac32$: the lemma does not apply there
yet, because orders $2,3$ are still uncancelled. Run the sweep: read $c=1$ at $k=2$
and add $\log(1-q^2)=-(q^2+\tfrac12 q^4+\cdots)$; read $c=1$ at $k=3$ and add
$\log(1-q^3)$. Now
\[
[q^4]R=\tfrac32-\tfrac12+0=1\in\mathbb{Z},
\]
exactly the next factor (power $4$, coefficient $+1$). The coefficient that was a
fraction in the raw log becomes a clean integer the moment the sweep is entitled to
read it. (The FORM demo \texttt{log\_expand.frm} prints exactly this: raw $L$ carries
$\tfrac32 q^4$, the residual after peeling orders $2,3$ carries $q^4$.)

\subsection{Cost}

Per factor at power $k$, updating $R$ to working order $N$ touches only the
$\lfloor N/k\rfloor$ multiples of $k$ --- an $O(N/k)$ sparse addition, versus the
$O(N)$ convolution the direct method needs. The only real overhead is arithmetic
over $\mathbb{Q}$ rather than $\mathbb{Z}$; since each read is an integer
(Lemma~\ref{lem:integer}), one can even reduce the tail modulo the processed part to
keep denominators small, though that is an optimization, not a necessity. In
practice the truncated-stream implementation produces 50 factors in a blink where
the dense FORM expansion hangs.

\section{The actual target: $1/\varphi(n)^2$}
\label{sec:target}

\begin{proposition}
\label{prop:bell-1overphi2}
The Bell factor of $1/\varphi(n)^2$ at prime $p$ is
\[
B_p = \frac{(p-1)^2p^2 + (2p-1)\,q^{s}}{(p-1)^2(p^2-q^{s})}, \qquad q = p^{-1}.
\]
\end{proposition}

\begin{proof}
Master equation: $a(p^e)=1/\varphi(p^e)^2=(p-1)^{-2}p^{-2e+2}$ for $e\ge1$,
$a(1)=1$. Then, with $q^{s}=p^{-s}$,
\[
B_p = 1 + \sum_{e\ge1}\frac{p^{2-2e}}{(p-1)^2}q^{se}
= 1 + \frac{p^2}{(p-1)^2}\cdot\frac{q^{s}/p^2}{1-q^{s}/p^2}
= 1 + \frac{q^{s}}{(p-1)^2}\cdot\frac{1}{1-q^{s}/p^2},
\]
and clearing denominators, with $p^2-(p-1)^2=2p-1$,
\[
B_p=\frac{(p-1)^2(p^2-q^{s})+q^{s}p^2}{(p-1)^2(p^2-q^{s})}
      =\frac{(p-1)^2p^2+(2p-1)q^{s}}{(p-1)^2(p^2-q^{s})}.
\]
\end{proof}

Normalizing to $B_p\to1$ as $q\to0$ (divide through by $(p-1)^2p^2$),
\begin{equation}
\label{eq:phi-inv-sq-normalized}
B_p = \frac{1+\dfrac{2p-1}{(p-1)^2p^2}\,q^{s}}{1-q^{\,s+2}}.
\end{equation}
The denominator $1-q^{s+2}$ is exactly $1-q^{\,us-l}$ with $(l,u)=(-2,1)$; sitting in
the denominator of $B_p$, it hands us a clean $\zeta(s+2)$ straight away (by
\eqref{eq:euler-minus} with $\gamma=-1$). The numerator is the problem: its
coefficient $\frac{2p-1}{(p-1)^2p^2}$ is not a Laurent monomial in $p$ (it is a
genuine rational function of $p$), so it is not of the $p^{c}$ shape the extraction
pattern-matches on. Two ways out:

\begin{itemize}
\item expand that coefficient as a series in $1/p$,
$\frac{2p-1}{(p-1)^2p^2}=\sum_k c_k p^{-k}$, and let each term seed its own zeta
factor --- so the ``numerator'' is really a full power series in $q$ whose
$p$-dependence at each order must be decomposed into $p$-power monomials; or
\item there is no finite factorization and the extraction is a genuinely infinite,
non-terminating recursion, whose exponent pattern we would conjecture numerically
(as Mathar did for $\varphi^2$) rather than derive in closed form.
\end{itemize}
Which one holds is a computation, not a guess: run the extractor to enough orders,
look for a recognizable exponent sequence (ideally cross-checked against OEIS, the
way Mathar does), and only then commit to a statement.

\section{Where this is going}

\begin{enumerate}
\item Run the exact-rational extractor first on Mathar's $\varphi^2(n)$
\eqref{eq:mathar-phi2} (16 known terms) and the terminating Jordan case
$J_k(n)/n^s\mapsto\zeta(s-k)/\zeta(s)$ as sanity checks.
\item Then run it on Proposition~\ref{prop:bell-1overphi2}'s $B_p$ for
$1/\varphi(n)^2$ to good order and extract/conjecture the $(l_i,u_i,\gamma_i)$
sequence.
\item Search OEIS for that sequence --- cheap and effective.
\item Only after (2)--(3) succeed, write the closed-form (or asymptotic
infinite-product) statement for $1/\varphi(n)^2$ as a theorem with proof.
\end{enumerate}

\begin{thebibliography}{9}
\bibitem{Mathar2011}
R.~J.~Mathar, \emph{Survey of Dirichlet Series of Multiplicative Arithmetic
Functions}, arXiv:1106.4038 [math.NT].
\end{thebibliography}

\end{document}
